---
title: "analyze word associations"
author: "Julius Fenn"
format:
  html:
    toc: true
    toc-depth: 4
    html-math-method: katex
---


# Notes

```{r}
## global variables: 
# only if necessary to limit running time
```



# load needed pre-processed data


```{r}
#| echo: true
#| warning: false

# sets the directory of location of this script as the current directory
# setwd(dirname(rstudioapi::getSourceEditorContext()$path))

# load packages
require(pacman)
p_load('xlsx', '', 'DT', 'igraph', 'tidyverse', 'wordcloud')


# load data
setwd("../generated data")

# for first order associations example
associationData_firstOrder <- xlsx::read.xlsx2(file = "word_associations_firstOrder_adjectives.xlsx", sheetIndex = 1)

## adjust ID variable
associationData_firstOrder$participant_id <- str_extract(string = associationData_firstOrder$participant_id, pattern = "[:alpha:]*_[:digit:]*")
```



# first order associations

## describe data

```{r}
dim(associationData_firstOrder)
table(associationData_firstOrder$cue)
length(unique(associationData_firstOrder$participant_id))
```

dynamic data table:

```{r}
DT::datatable(data = associationData_firstOrder)
```

## analyze data

### get word clouds


function to create word clouds: 

```{r}
generate_wordcloud <- function(cue_word, data_set) {
  
  # Filter the data set for the given cue word
  tmp <- data_set[data_set$cue == cue_word, ]
  
  # Calculate word frequencies
  word_frequencies <- table(tmp$response)
  word_frequencies_df <- as.data.frame(word_frequencies)
  
  # Create the word cloud
  wordcloud(words = word_frequencies_df$Var1, 
            freq = word_frequencies_df$Freq, 
            min.freq = 1, 
            max.words = 100, 
            random.order = FALSE, 
            colors = brewer.pal(8, "Dark2"))
}
```



```{r}
generate_wordcloud("muskulös", associationData_firstOrder)
```

```{r}
generate_wordcloud("normalgewichtig", associationData_firstOrder)
```


```{r}
generate_wordcloud("übergewichtig", associationData_firstOrder)
```


```{r}
generate_wordcloud("untergewichtig", associationData_firstOrder)
```


### get semantic networks

The idea of Network Inference from Fluency Data is motivated by two articles:

* Wulff, D. U., Hills, T. T., & Mata, R. (2022). Structural differences in the semantic networks of younger and older adults. Scientific Reports, 12(1), Article 1. https://doi.org/10.1038/s41598-022-11698-4
* Goñi, J., Arrondo, G., Sepulcre, J., Martincorena, I., Vélez de Mendizábal, N., Corominas-Murtra, B., Bejarano, B., Ardanza-Trevijano, S., Peraita, H., Wall, D. P., & Villoslada, P. (2011). The semantic organization of the animal category: Evidence from semantic verbal fluency and network theory. Cognitive Processing, 12(2), 183–196. https://doi.org/10.1007/s10339-010-0372-x



function to create semantic networks (! without checking if all the edges are "true", appear more frequently than expected randomly): 

```{r}
  # Function to create edges from the grouped responses
  create_edges <- function(association, distance = 2) {
    edges <- data.frame()
    n <- length(association)
    
    for (i in 1:n) {
      for (d in 1:distance) {
        if (i + d <= n) {
          edges <- rbind(edges, data.frame(from = association[i], to = association[i + d]))
        }
      }
    }
    return(edges)
  }

generate_association_network <- function(cue_word, data_set, distance = 2) {
  # Filter the dataset based on the cue word
  tmp <- data_set[data_set$cue == cue_word, ]
  
  # print number of times a response was given:
  print("number of times a response was given")
  print(sort(table(tmp$response)))
  
  grouped_responses <- tmp %>%
    group_by(participant_id) %>%
    summarise(associations = list(response)) %>%
    ungroup()
  
  # Apply the function to create edges for each participant
  all_edges <- do.call(rbind, lapply(grouped_responses$associations, create_edges, distance = distance))
  
  # Calculate word frequencies across the entire data set
  word_frequencies <- tmp %>%
    count(response, name = "frequency")
  
  # Calculate the edge weights (how often each link appears between two words)
  edge_weights <- all_edges %>%
    group_by(from, to) %>%
    summarise(weight = n(), .groups = "drop")
  
  # Create the graph from the edge list with edge weights
  g <- graph_from_data_frame(edge_weights, directed = FALSE)
  
  # Get the list of unique words in the graph (vertices)
  vertices <- V(g)$name
  
  # Extract edge weights to adjust the edge widths
  edge_widths <- E(g)$weight
  
  # Map word frequencies to vertex sizes
  vertex_sizes <- sapply(vertices, function(word) {
    freq <- word_frequencies %>%
      filter(response == word) %>%
      pull(frequency)
    

    if(length(freq) == 0) {
      return(1)  # Default size if the word is not found (shouldn't happen)
    } else {
      return(freq)
    }
  })
  
  # Plot the network with vertex sizes proportional to word frequencies and edge widths proportional to edge weights
  plot(g, 
       vertex.label.color = "black", 
       vertex.size = vertex_sizes * 1.5,           # Scale vertex sizes by word frequency
       edge.width = edge_widths * 1,             # Scale edge widths by weight (number of links)
       edge.arrow.size = 0.5)
  
  # Return the igraph object
  return(g)
}
```





```{r}
tmp_g <- generate_association_network("muskulös", associationData_firstOrder, distance = 2)

mat <- as.matrix(as_adjacency_matrix(tmp_g, attr = "weight"))

define_threshold <- 10
mat[apply(mat, 1, function(row) any(row >= define_threshold)), apply(mat, 2, function(row) any(row >= define_threshold))]
```


```{r}
tmp_g <- generate_association_network("normalgewichtig", associationData_firstOrder, distance = 2)

mat <- as.matrix(as_adjacency_matrix(tmp_g, attr = "weight"))

define_threshold <- 10
mat[apply(mat, 1, function(row) any(row >= define_threshold)), apply(mat, 2, function(row) any(row >= define_threshold))]
```


```{r}
tmp_g <- generate_association_network("übergewichtig", associationData_firstOrder, distance = 2)

mat <- as.matrix(as_adjacency_matrix(tmp_g, attr = "weight"))

define_threshold <- 6
mat[apply(mat, 1, function(row) any(row >= define_threshold)), apply(mat, 2, function(row) any(row >= define_threshold))]
```


```{r}
tmp_g <- generate_association_network("untergewichtig", associationData_firstOrder, distance = 2)

mat <- as.matrix(as_adjacency_matrix(tmp_g, attr = "weight"))

define_threshold <- 6
mat[apply(mat, 1, function(row) any(row >= define_threshold)), apply(mat, 2, function(row) any(row >= define_threshold))]
```





# second order associations (not shown)

just some random code from the associatoR package


```{r}
#| eval: false


ar_obj <- ar_import(associationData_firstOrder,
          participant = participant_id,
          cue = cue,
          response = response,
          response_vars = c(response_position))
str(ar_obj)

# normalize responses
ar_obj <- ar_normalize(ar_obj,
                       case = "most_frequent",
                       punct = "all",
                       whitespace = "squish",
                       process_cues = TRUE)


ar_obj <- ar_set_targets(ar_obj, targets = "cues")


ar_obj <- ar_summarize_targets(ar_obj,
                               response_var = response_position,
                               fun = mean)



ar_obj$targets


ar_obj <- ar_embed_targets(ar_obj,
                           method = "ppmi-svd",
                           n_dim = 100)

projection <- ar_project_embedding(ar_obj, method = "umap")



ar_plot_wordcloud(
  ar_obj,
  facet_col = cues,
  facet_row = NULL,
  color_by = cues,
  top_n = 20
)




# Importing the associationData_firstOrder data
ar_obj <- ar_import(data = associationData_firstOrder,
                    participant = participant_id,
                    cue = cue,
                    response = response,
                    participant_vars = c(gender,
                                         agegroup),
                    response_vars = c(response_position))




# export wordlist
# associatoR::ar_wordlist_export(ar_obj,
#                    file = "outputs/associationData_firstOrder_wordlist.xlsx")
# import wordlist
# ar_obj <- associatoR::ar_wordlist_import(ar_obj,
#                              file = "associationData_firstOrder_wordlist.csv",
#                              process_cues = TRUE,
#                              na = c("NA"))




# ar_obj <- ar_count_targets(ar_obj) # to get frequencies of target
# ar_obj$targets

ar_obj <- ar_characterize_targets(
  ar_obj,
  characteristics = c("valence", "arousal", "dominance"),
  case_sensitive = FALSE
) # "word_frequency", "concreteness"
ar_obj$targets


ar_obj <- ar_summarize_targets(ar_obj,
                               response_var = response_position,
                               fun = mean)
ar_obj$targets



ar_obj <- ar_correlate_targets(
  ar_obj,
  participant_vars = c(gender, agegroup),
  metric = "auto"
)

ar_obj$targets$target[order(ar_obj$targets$gender_corr)[1:5]] # more frequently by women
ar_obj$targets$target[order(ar_obj$targets$gender_corr, decreasing = TRUE)[1:5]] # more frequently by man


#######################################
tmp <- ar_cross_targets(ar_obj,
                        participant_vars = c(gender), # agegroup
                        target_var = target,
                        normalize = FALSE)
tmp[tmp$target == "empathy",]
tmp[tmp$target == "Elon Musk",]
#######################################


ar_obj <- ar_cluster_targets(ar_obj,
                             method = "louvain",
                             similarity = "cosine",
                             resolution = 1.1)
table(ar_obj$targets$cluster)


# ar_plot_embedding(
#   ar_obj,
#   color_by = cluster,
#   color_set = "G",
#   alpha = 0.5,
#   proportion_labels = 1
# )

ar_import(associationData_firstOrder,
          participant = participant_id,
          cue = cue,
          response = response,
          participant_vars = c(gender, agegroup),
          response_vars = c(response_position, response_level)) %>%
  ar_set_targets("cues") %>%
  ar_embed_targets() %>%
  ar_cluster_targets() %>%
  ar_project_embedding() %>%
  ar_plot_embedding(color_by = cluster,
                    proportion_labels = .5)



ar_plot_wordcloud(
  ar_obj,
  facet_col = cluster,
  facet_row = NULL,
  color_by = cluster,
  top_n = 20
)


# ar_cluster_stability(ar_obj)



ar_cross_targets(ar_obj,
                participant_vars = c(gender), # agegroup
                target_var = cluster,
                normalize = TRUE)


ar_compare_targets(ar_obj,
                   participant_vars = c(gender, agegroup),
                   target_var = valence,
                   fun = median, na.rm = TRUE)



ar_compare_embeddings(ar_obj,
                      participant_vars = c(gender, agegroup),
                      type = "triangle",
                      intersection = "pair")


ar_import(associationData_firstOrder,
          participant = participant_id,
          cue = cue,
          response = response,
          response_vars = c(response_position)) %>%
  ar_set_targets("cues") %>%
  ar_embed_targets(method = "counts") %>%
  ar_cluster_targets(method = "hclust", k = 3) %>%
  ar_project_embedding(method = "mds") %>%
  ar_plot_embedding(color_by = cluster,
                    proportion_labels = .5)



ar_plot_wordcloud(
  ar_obj,
  facet_col = cluster,
  facet_row = NULL,
  color_by = cluster,
  top_n = 20
)
```



